---
title: "Working with cohorts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  # eval = FALSE,
  comment = "#>"
)

library(CDMConnector)
library(dplyr, warn.conflicts = FALSE)

if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = file.path(tempdir(), "eunomia"))
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!eunomia_is_available()) downloadEunomiaData()

```


## Adding a cohort

First, we'll load packages and create a cdm reference. In this case we'll be using the Eunomia "GI Bleed" dataset.

```{r, warning=FALSE, message=FALSE}
library(CDMConnector)
library(dplyr)

con <- DBI::dbConnect(duckdb::duckdb(), eunomia_dir())

cdm <- CDMConnector::cdm_from_con(
  con = con,
  cdm_schema = "main",
  write_schema = "main"
)
```

We can define a cohort for GI bleeding, where we exclude anyone with a record of rheumatoid arthritis at any time.

```{r, eval=FALSE}
# devtools::install_github("OHDSI/Capr")
library(Capr)

gibleed_cohort_definition <- cohort(
  entry = condition(cs(descendants(192671))),
  attrition = attrition(
    "no RA" = withAll(
      exactly(0,
              condition(cs(descendants(80809))),
              duringInterval(eventStarts(-Inf, Inf))))
  )
)

# requires CirceR optional dependency
cdm <- generateCohortSet(
  cdm,
  cohortSet = list(gibleed = gibleed_cohort_definition),
  name = "gibleed",
  computeAttrition = TRUE
)
```

```{r, echo=FALSE, eval=FALSE}
# save data in package. Only do this when editing vignette.
cdm$gibleed %>% 
  collect() %>% 
  readr::write_rds(here::here("inst", "rds", "gibleed.rds"))

cdm$gibleed %>% 
  cohortCount() %>% 
  collect() %>% 
  readr::write_rds(here::here("inst", "rds", "gibleed_count.rds"))

cdm$gibleed %>% 
  cohortAttrition() %>% 
  collect() %>% 
  readr::write_rds(here::here("inst", "rds", "gibleed_attrition.rds"))

cdm$gibleed %>% 
  cohortSet() %>% 
  collect() %>% 
  readr::write_rds(here::here("inst", "rds", "gibleed_set.rds"))

```


We can see that we now have our cohort instantiated in the database with a reference to it added to the cdm reference.
```{r, eval=FALSE}
cdm$gibleed %>% 
  glimpse()
```
```{r, echo=FALSE}
readr::read_rds(system.file("rds", "gibleed.rds", package = "CDMConnector", mustWork = TRUE)) %>% 
  glimpse()
```


## Cohort attributes

As well as the cohort itself, the cohort has a number of attributes. First, is a count of participants by cohort. We can use `cohortCount` to get these counts.

```{r, eval=FALSE}
cohortCount(cdm$gibleed) %>% 
  glimpse()
```

```{r, echo=FALSE}
readr::read_rds(system.file("rds", "gibleed_count.rds", package = "CDMConnector", mustWork = TRUE)) %>% 
  glimpse()
```

We also have the attrition associated with entry into the cohort available via `cohortAttrition`.
```{r, eval=FALSE}
cohortAttrition(cdm$gibleed) %>% 
  glimpse()
```
```{r, echo=FALSE}
readr::read_rds(system.file("rds", "gibleed_attrition.rds", package = "CDMConnector", mustWork = TRUE)) %>% 
  glimpse()
```

And lastly, we can also access the settings associated with the cohort using `cohortCount`.
```{r, eval=FALSE}
cohortSet(cdm$gibleed) %>% 
  glimpse()
```

```{r, echo=FALSE}
readr::read_rds(system.file("rds", "gibleed_set.rds", package = "CDMConnector", mustWork = TRUE)) %>% 
  glimpse()
```

```{r}
DBI::dbDisconnect(con, shutdown = TRUE)
```

