---
title: "Using CDM attributes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using CDM attributes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  # eval = FALSE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(CDMConnector)
if (Sys.getenv("EUNOMIA_DATA_FOLDER") == "") Sys.setenv("EUNOMIA_DATA_FOLDER" = file.path(tempdir(), "eunomia"))
if (!dir.exists(Sys.getenv("EUNOMIA_DATA_FOLDER"))) dir.create(Sys.getenv("EUNOMIA_DATA_FOLDER"))
if (!eunomia_is_available()) downloadEunomiaData()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Set up
Let's again load required packages and connect to our Eunomia dataset in duckdb.

```{r, message=FALSE, warning=FALSE}
library(CDMConnector)
library(omopgenerics)
library(dplyr)

write_schema <- "main"
cdm_schema <- "main"

con <- DBI::dbConnect(duckdb::duckdb(), dbdir = eunomia_dir())
cdm <- cdm_from_con(con, cdm_name = "eunomia", cdm_schema = cdm_schema, write_schema = write_schema)
```

## CDM reference attributes

Our cdm reference has various attributes associated with it. These can be useful both when programming and when developing analytic packages on top of CDMConnector.  

### CDM name
It's a requirement that every cdm reference has name associated with it. This is particularly useful for network studies so that we can associate results with a particular cdm. We can access this attribute like so 
```{r}
#attr(cdm, "cdm_name")
```

Because it is so regularly used, to make getting the cdm name even easier, we can also use `cdmName` (or it's snake case equivalent `cdm_name`)
```{r}
cdmName(cdm)
cdm_name(cdm)
```

### CDM version
The OMOP CDM has various versions. We also have an attribute giving the version of the cdm we have connected to.
```{r}
#attr(cdm, "cdm_version")
```

### Database connection
We also have an attribute identifying the database connection underlying the cdm reference.
```{r}
#cdmCon(cdm)
```

This can be useful, for example, if we want to make use of DBI functions to work with the database. For example we could use `dbListTables` to list the names of remote tables accessible through the connection, `dbListFields` to list the field names of a specific remote table, and `dbGetQuery` to returns the result of a query 
```{r}
#DBI::dbListTables(cdmCon(cdm))
#DBI::dbListFields(cdmCon(cdm), "person")
#DBI::dbGetQuery(cdmCon(cdm), "SELECT * FROM person LIMIT 5")
```

## Cohort attributes
### Generated cohort set
When we generate a cohort in addition to the cohort table itself we also have various attributes that can be useful for subsequent analysis.

Here we create a cohort table with a single cohort.
```{r}

# debugonce(generateConceptCohortSet)
cdm <- generateConceptCohortSet(cdm = cdm, 
                                conceptSet = list("gi_bleed" = 192671,
                                                  "celecoxib" = 1118084), 
                                name = "study_cohorts",
                                overwrite = TRUE)

cdm$study_cohorts %>% 
  glimpse()
```

We have a cohort set attribute that gives details on the settings associated with the cohorts (along with utility functions to make it easier to access this attribute).
```{r}
attr(cdm$study_cohorts, "cohort_set")
```

```{r, eval=FALSE}
cohortSet(cdm$study_cohorts)
cohort_set(cdm$study_cohorts) 
```

We have a cohort_count attribute with counts for each of the cohorts.
```{r}
attr(cdm$study_cohorts, "cohort_count")
```

```{r, eval=FALSE}
cohortCount(cdm$study_cohorts)
cohort_count(cdm$study_cohorts)
```

And we also have an attribute, cohort attrition, with a summary of attrition when creating the cohorts.

```{r, eval=FALSE}
attr(cdm$study_cohorts, "cohort_attrition")
```

```{r, eval=FALSE}
cohortAttrition(cdm$study_cohorts)
cohort_attrition(cdm$study_cohorts)
```

In addition, we also have the cdm reference itself as an attribute of the cohorts. This is particularly useful when developing analytic packages on top of CDMConnector.
```{r}
attr(cdm$study_cohorts, "cdm_reference")
```

### Creating a bespoke cohort
Say we create a custom GI bleed cohort with the standard cohort structure
```{r}
cdm$gi_bleed <- cdm$condition_occurrence %>% 
  filter(condition_concept_id == 192671) %>% 
  mutate(cohort_definition_id = 1) %>% 
  select(
    cohort_definition_id, 
    subject_id = person_id, 
    cohort_start_date = condition_start_date, 
    cohort_end_date = condition_start_date
  ) %>% 
  compute(name = "gi_bleed", temporary = FALSE, overwrite = TRUE)

cdm$gi_bleed %>% 
  glimpse()
```

We can add the required attributes using the `newGeneratedCohortSet` function. The minimum requirement for this is that we also define the cohort set to associate with our set of custom cohorts.

```{r}
GI_bleed_cohort_ref <- tibble(cohort_definition_id = 1, cohort_name = "custom_gi_bleed")

cdm$gi_bleed <- omopgenerics::cohortTable(
  table = cdm$gi_bleed, cohortSetRef = GI_bleed_cohort_ref
)
```

Now our custom cohort GI_bleed has the same attributes associated with it as if it had been created by `generateConceptCohortSet`. This will mean that it can be used by analytic packages designed to work with cdm cohorts.

```{r}
cohortSet(cdm$gi_bleed)
cohortCount(cdm$gi_bleed)
cohortAttrition(cdm$gi_bleed)
attr(cdm$gi_bleed, "cdm_reference")
```


<div style="margin-bottom:3cm;"></div>
