---
title: "Using SQL from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using SQL from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```


It is common to send SQL to an OMOP CDM database from R. This vignette describes one process for including SQL in your analytic package or study.

```{r setup}
library(CDMConnector)
```


If your study needs to run across multiple database platforms then you will need a way to translate SQL into the different dialects used by each database. The `SqlRender` package provides this functionality. Start by creating an R function that includes the SQL you want to run and takes a database connection as the first argument. Additional query parameters should be added as detail arguments after the first one.

```{r}
count_drug_records <- function(con, cdm_schema, ingredient_concept_id) {
  
  # Write the SQL in an R function replacing the parameters
  sql <- glue::glue("
    select count(*) as drug_count 
    from {cdm_schema}.drug_exposure de
    inner join {cdm_schema}.concept_ancestor ca
    on de.drug_concept_id = ca.descendant_concept_id
    where ca.ancestor_concept_id = {ingredient_concept_id}")
  
  # Translate the sql to the correct dialect
  sql <- SqlRender::translate(sql, targetDialect = dbms(con))
  
  # Execute the SQL and return the result
  DBI::dbGetQuery(con, sql)
}
```



```{r, eval=FALSE}
library(DBI)
con <- DBI::dbConnect(RPostgres::Postgres(),
                      dbname = "cdm",
                      host = "localhost",
                      user = "postgres",
                      password = Sys.getenv("password"))

count_drug_records(con, cdm_schema = "synthea1k", ingredient_concept_id = 923672)
```

```{r, echo=FALSE}
dplyr::tibble(drug_count = 11)
```

The CDMConnector package has a function `dbms` that will return the correct dialect for a connection.

```{r, eval=FALSE}
dbms(con)
```

```{r, echo=FALSE}
print(c("postgresql"))
```

`SQLRender` also includes a function called `render` to insert values into parameterized SQL.

```{r, eval=FALSE}
count_drug_records <- function(con, cdm_schema, ingredient_concept_id) {
  
  # Write the SQL in an R function replacing the parameters
  sql <- SqlRender::render("
    select count(*) as drug_count 
    from @cdm_schema.drug_exposure de
    inner join @cdm_schema.concept_ancestor ca
    on de.drug_concept_id = ca.descendant_concept_id
    where ca.ancestor_concept_id = @ingredient_concept_id",
    cdm_schema = cdm_schema, 
    ingredient_concept_id = ingredient_concept_id)
  
  # Translate the sql to the correct dialect
  sql <- SqlRender::translate(sql, targetDialect = dbms(con))
  
  # Execute the SQL and return the result
  DBI::dbGetQuery(con, sql)
}

count_drug_records(con, cdm_schema = "synthea1k", ingredient_concept_id = 923672)
```

```{r, echo=FALSE}
dplyr::tibble(drug_count = 11)
```


`SqlRender` translates a particular dialect of SQL known as OHDSI-SQL to various supported database platforms. OHDSI-SQL is loosely defined as a subset of Microsoft T-SQL used by SQL Server. It also has branching logic and parameterization added to it. 

More details about using SqlRender to run SQL across multiple database platforms can be found in the "SQL and R" chapter of the ["The Book of OHDSI"](https://ohdsi.github.io/TheBookOfOhdsi/SqlAndR.html).

The [OHDSI Query Library](https://data.ohdsi.org/QueryLibrary/) contains example SQL queries for the OMOP CDM.

<div style="margin-bottom:3cm;"></div>
