---
title: "Working with cohorts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working_with_cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

Sys.setenv(EUNOMIA_DATA_FOLDER = tempdir())
```


## Adding a cohort

First, we'll load packages and create a cdm reference. In this case we'll be using the Eunomia "GI Bleed" dataset.

```{r, warning=FALSE, message=FALSE, eval=TRUE}
library(CDMConnector)
library(dplyr)

con <- DBI::dbConnect(duckdb::duckdb(), eunomia_dir())

cdm <- CDMConnector::cdm_from_con(
  con = con,
  cdm_schema = "main",
  write_schema = "main"
)
```

We can define a cohort for GI bleeding, where we exclude anyone with a record of rheumatoid arthritis at any time.

```{r, eval=FALSE}
# devtools::install_github("OHDSI/Capr", "v2") # Use the development version of Capr v2
library(Capr)

# create a cohort set folder for saving cohort definitions
path <- file.path(tempdir(), "cohorts")
dir.create(path)

gibleed_cohort_definition <- cohort(
  entry = condition(cs(descendants(192671))),
  attrition = attrition(
    "no RA" = withAll(
      exactly(0,
              condition(cs(descendants(80809))),
              duringInterval(eventStarts(-Inf, Inf))))
  )
)

# write json cohort definition
writeCohort(gibleed_cohort_definition, file.path(path, "gibleed.json"))
```

```{r, echo=FALSE}
# use internal cohort set
gibleed_cohort_set <- readCohortSet(system.file("cohorts2", package = "CDMConnector", mustWork = TRUE))[3,]
```

Now we have our cohort definition, we can generate the cohort.

```{r, eval=FALSE}
gibleed_cohort_set <- readCohortSet(path = path)
```

```{r}
cdm <- generateCohortSet(
  cdm,
  gibleed_cohort_set,
  name = "gibleed",
  computeAttrition = TRUE
)
```

We can see that we now have our cohort instantiated in the database with a reference to it added to the cdm reference.
```{r}
cdm$gibleed %>% 
  glimpse()
```

## Cohort attributes

As well as the cohort itself, the cohort has a number of attributes. First, is a count of participants by cohort. We can use `cohortCount` to get these counts.

```{r}
cohortCount(cdm$gibleed)
```

We also have the attrition associated with entry into the cohort available via `cohortAttrition`.
```{r}
cohortAttrition(cdm$gibleed)
```

And lastly, we can also access the settings associated with the cohort using `cohortCount`.
```{r}
cohortSet(cdm$gibleed)
```

```{r}
DBI::dbDisconnect(con, shutdown = TRUE)
```

