---
title: "How to Test Study Code on Target Databases"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Overview

When developing an OHDSI study package you often want to verify that your
analysis code runs correctly on the same database platform your collaborators
use (PostgreSQL, SQL Server, Databricks/Spark, etc.). The workflow below shows
how to:

1. Create a test CDM locally with `TestGenerator::patientsCDM()`
2. Trim the vocabulary to the minimum needed with `cdmTrimVocabulary()`
3. Upload the test CDM to a remote database with `copyCdmTo()`
4. Run your study on the remote CDM
5. Clean up the test schema

## 1. Load test data

Use `TestGenerator::patientsCDM()` to build a small CDM backed by DuckDB.

```{r generate}
# this code requires development version of CDMConnector
remotes::install_github("darwin-eu-dev/CDMConnector", upgrade = "never")

library(CDMConnector)
cdm <- TestGenerator::patientsCDM("~/darwin_studies/AML/P4-C1-007/inst/testPatients/aml10")

cdm$person
```

## 2. Trim the vocabulary

A full OMOP vocabulary has millions of rows. For a small test CDM most of that
is unnecessary and slows down the upload. `cdmTrimVocabulary()` keeps only the
concepts that actually appear in the CDM plus their ancestors in
`concept_ancestor` (needed for concept set resolution). Tables like
`concept_synonym` and `source_to_concept_map` are emptied entirely.

```{r trim}
cdm <- cdmTrimVocabulary(cdm)
```

After trimming you can inspect the size:

```{r check-size}
cdm$concept |> dplyr::tally()
cdm$concept_ancestor |> dplyr::tally()
cdm$concept_relationship |> dplyr::tally()
```

## 3. Connect to a remote database

Create a DBI connection and choose a dedicated schema (or prefix) for the test
data. Using a separate schema makes cleanup easy: you can drop the entire
schema when you are done.

### PostgreSQL

```{r pg-connect}
con <- DBI::dbConnect(
  RPostgres::Postgres(),
  dbname   = Sys.getenv("CDM5_POSTGRESQL_DBNAME"),
  host     = Sys.getenv("CDM5_POSTGRESQL_HOST"),
  user     = Sys.getenv("CDM5_POSTGRESQL_USER"),
  password = Sys.getenv("CDM5_POSTGRESQL_PASSWORD")
)

# Create a dedicated schema for the test CDM
test_schema <- "test_study_cdm"
DBI::dbExecute(con, paste("CREATE SCHEMA IF NOT EXISTS", test_schema))
```

### SQL Server

```{r sqlserver-connect}
con <- DBI::dbConnect(
  odbc::odbc(),
  Driver   = Sys.getenv("SQL_SERVER_DRIVER"),
  Server   = Sys.getenv("CDM5_SQL_SERVER_SERVER"),
  Database = Sys.getenv("CDM5_SQL_SERVER_CDM_DATABASE"),
  UID      = Sys.getenv("CDM5_SQL_SERVER_USER"),
  PWD      = Sys.getenv("CDM5_SQL_SERVER_PASSWORD"),
  TrustServerCertificate = "yes",
  Port     = Sys.getenv("CDM5_SQL_SERVER_PORT")
)

# Create a dedicated schema for the test CDM
test_schema <- "test_study_cdm"
DBI::dbExecute(con, paste("CREATE SCHEMA", test_schema))
```

### Databricks / Spark

```{r spark-connect}
con <- DBI::dbConnect(
  odbc::databricks(),
  httpPath       = Sys.getenv("DATABRICKS_HTTPPATH"),
  useNativeQuery = FALSE,
  bigint         = "numeric"
)

# On Databricks, schemas live inside a catalog
test_schema <- c(
  catalog = Sys.getenv("DATABRICKS_CATALOG", "hive_metastore"),
  schema  = "test_study_cdm"
)

DBI::dbExecute(con, paste0(
  "CREATE SCHEMA IF NOT EXISTS ",
  paste(test_schema, collapse = ".")
))
```

## 4. Upload the test CDM

`copyCdmTo()` collects each table from the local DuckDB CDM and inserts it
into the remote schema. Because we trimmed the vocabulary the upload is fast.

```{r upload}
remote_cdm <- copyCdmTo(
  con       = con,
  cdm       = cdm,
  schema    = test_schema,
  overwrite = TRUE
)
```

## 5. Run your study

Now you can run your study code against the remote CDM exactly as a
collaborator would. Here is a minimal example:

```{r run-study}
# A simple study function for demonstration purposes
runStudy <- function(cdm) {
  message("Starting study execution on: ", omopgenerics::cdmName(cdm))
  message("Database platform: ", CDMConnector::dbms(cdm))

  # Check the CDM
  snap <- CDMConnector::snapshot(cdm)
  message("Persons: ", snap$person_count)
  message("Vocabulary version: ", snap$vocabulary_version)

  # -------------------------------------------------------
  # Your real study code goes here, e.g.:
  #   results <- CohortIncidence::executeStudy(cdm, ...)
  #   results <- CohortCharacteristics::summarise(cdm, ...)
  # -------------------------------------------------------

  message("Study execution complete.")
  return(snap)
}

results <- runStudy(remote_cdm)
results
```

## 6. Clean up

After you have verified that the study runs, drop the test schema to free
space. **Be careful** -- this permanently deletes all tables in the schema.

### PostgreSQL

```{r cleanup-pg}
DBI::dbExecute(con, paste("DROP SCHEMA", test_schema, "CASCADE"))
DBI::dbDisconnect(con)
```

### SQL Server

```{r cleanup-sqlserver}
# SQL Server requires dropping all tables before the schema
tables <- DBI::dbListTables(con)
for (tbl in tables) {
  DBI::dbExecute(con, paste0("DROP TABLE IF EXISTS ", test_schema, ".", tbl))
}
DBI::dbExecute(con, paste("DROP SCHEMA", test_schema))
DBI::dbDisconnect(con)
```

### Databricks / Spark

```{r cleanup-spark}
DBI::dbExecute(con, paste0(
  "DROP SCHEMA IF EXISTS ",
  paste(test_schema, collapse = "."),
  " CASCADE"
))
DBI::dbDisconnect(con)
```

## Summary

| Step | Function | Purpose |
|------|----------|---------|
| Create CDM | `TestGenerator::patientsCDM` | Create test CDM locally |
| Trim | `cdmTrimVocabulary()` | Remove unused vocabulary rows |
| Connect | `DBI::dbConnect()` | Open connection to target platform |
| Upload | `copyCdmTo()` | Copy test CDM to remote schema |
| Run | `runStudy(cdm)` | Execute study code on remote data |
| Clean up | `DROP SCHEMA ... CASCADE` | Remove test data from remote database |
